{\Large \textbf{Abstract}}

\bigskip

Recreational and competitive athletes rely on multiple digital platforms to track training load, monitor recovery, plan routes, and schedule sessions. However, this data remains fragmented across proprietary ecosystems---Strava for activities, Garmin for wellness, weather services for planning, and calendars for availability---forcing users into manual cross-referencing that is time-consuming and error-prone. This paper presents \textit{Training Copilot}, an open-source, multi-agent sports analytics platform that unifies these heterogeneous data sources behind a single conversational interface. Training Copilot adopts the Model Context Protocol (MCP) as a uniform integration layer: each external API is encapsulated in an independent MCP server, and a central host discovers and routes tool calls without hardcoding provider-specific logic. On top of this data layer, a LangGraph-based multi-agent system coordinates five specialist agents---recovery, training load, environmental context, route planning, and fitness knowledge---via the Agent-to-Agent (A2A) protocol. Each specialist is an autonomous Reasoning-and-Acting (ReAct) agent scoped to its own MCP servers. An orchestrator agent decomposes user queries, delegates to the appropriate specialists in parallel, and synthesises data-driven recommendations. The fitness knowledge specialist employs Retrieval-Augmented Generation (RAG) over a curated vector database of public-domain exercise science literature, grounding its advice in verifiable sources. We evaluate the system through an automated end-to-end assessment using MLflow's conversation simulation framework, employing ten synthetic personas across two user archetypes that collectively exercise all five specialist domains. Each conversation is scored along five dimensions---conversation completeness (whether the user's goal was achieved), user frustration (resilience under adversarial follow-ups), safety (absence of harmful advice), supportive coaching tone (empathy under pushback), and data grounding---where the last scorer is fully deterministic, verifying tool usage from execution traces rather than from chat text. The MCP-based architecture enables seamless extensibility---adding a new data source requires exactly one file and one configuration line---while the multi-agent design produces contextually aware, data-grounded training recommendations that integrate recovery status, workload trends, weather conditions, and route options into a single coherent response.
