\markboth{STATE OF THE ART}{STATE OF THE ART}
\section{State of the Art}\label{sec:sota}

This section reviews the current state of research across the four pillars that Training Copilot builds upon: agentic AI and multi-agent systems, tool integration protocols, sports analytics and wearable technology, and Retrieval-Augmented Generation.

\subsection{Agentic AI and Multi-Agent Systems}\label{sec:sota:agents}

The rapid scaling of large language models has enabled a paradigm shift from static text generation to autonomous \textit{agentic} systems that plan, reason, and interact with external environments. The Transformer architecture~\citep{vaswani_attention_2017}, chain-of-thought prompting~\citep{wei_chain-of-thought_2022}, and the scaling of model capacity to hundreds of billions of parameters~\citep{openai_gpt4_2023} laid the groundwork by demonstrating that LLMs can perform multi-step reasoning when prompted to generate intermediate steps. We organise the subsequent literature along three themes: reasoning architectures, agent-internal structure, and multi-agent coordination.

\subsubsection{Reasoning Architectures}

Two complementary patterns have emerged for grounding LLM reasoning in external action. \cite{yao_react_2023} introduced \textit{ReAct}, which interleaves reasoning traces (``thought'' steps) and task-specific actions in a single LLM loop: the model generates a thought explaining its plan, executes an action (e.g., \texttt{search[entity]} against a Wikipedia API), observes the result, and repeats until the task is complete. Evaluated on PaLM-540B across four benchmarks, ReAct achieved 71\,\% success rate on ALFWorld (vs.\ 45\,\% for action-only) and 40\,\% on WebShop (vs.\ 30.1\,\% for action-only). On knowledge tasks (HotpotQA, FEVER), ReAct produced zero hallucinated reasoning chains, compared to 56\,\% for chain-of-thought prompting---because every factual claim is grounded in a retrieved observation rather than generated from the model's parameters. Training Copilot's specialist agents use this pattern directly: each is a LangGraph ReAct agent where the ``actions'' are MCP tool calls.

\cite{shinn_reflexion_2023} extended this loop across episodes with \textit{Reflexion}, where agents store linguistic self-reflections after failed attempts in an episodic memory buffer, informing subsequent trials without weight updates. The two approaches are complementary: ReAct provides the within-episode reasoning loop, while Reflexion adds cross-episode learning. Training Copilot uses Reflexion in one specific subsystem: the LLM-driven chart generator retries once with the error message as feedback when execution fails (Section~\ref{sec:architecture:supporting}). Both patterns, however, assume a \textit{single} agent operating over a \textit{fixed} tool set---neither addresses how to coordinate multiple agents or how to discover tools dynamically from heterogeneous sources.

\subsubsection{Agent-Internal Architecture}

Several frameworks have sought to characterise the internal structure that makes an agent effective. \cite{park_generative_2023} demonstrated that LLM-powered agents exhibit believable behaviour when equipped with three components: a \textit{memory stream} recording experiences, a \textit{reflection} mechanism synthesising abstractions, and a \textit{planning} module translating reflections into action sequences. \cite{wang_survey_2024} generalised these ideas into a four-module taxonomy---profiling, memory, planning, and action---that subsumes prior architectures. A common finding across both works is that memory and planning are mutually reinforcing: richer memory enables better planning, which in turn generates more informative experiences. However, both frameworks describe single-agent architectures; they do not address how multiple agents with different profiles and tool sets should coordinate.

\subsubsection{Multi-Agent Coordination}

The transition to multi-agent systems introduces coordination challenges absent from single-agent designs. Three approaches illustrate the design space. \cite{shen_hugginggpt_2023} proposed HuggingGPT, where a central controller decomposes requests, selects specialist models from a hub, executes them, and summarises results---demonstrating the viability of orchestrator--specialist decomposition, though applied to model selection rather than tool-based data retrieval. \cite{wu_autogen_2023} introduced AutoGen, showing through empirical benchmarks that role-specialised agents with tool access outperform monolithic single-agent designs on complex tasks, and that flexible conversation patterns (sequential, parallel, nested) enable different coordination strategies. \cite{guo_large_2024} surveyed the field comprehensively and identified two open challenges that recur across systems: \textit{communication protocol design} (how agents exchange information without losing context) and \textit{task decomposition} (how to split a user request into specialist-sized subtasks). More broadly, \cite{acharya_agentic_2025} distinguished agentic AI from traditional automation by its capacity for autonomous goal pursuit through perception, reasoning, and action.

These works converge on a shared conclusion: multi-agent architectures with domain-scoped specialists outperform monolithic agents on tasks requiring heterogeneous knowledge. Yet they also share a limitation: none demonstrate integration with \textit{standardised} tool-discovery or inter-agent protocols, relying instead on ad-hoc tool registries and custom communication layers.

\subsection{Tool Integration Protocols}\label{sec:sota:protocols}

The practical utility of LLM agents depends on their ability to interact with external tools and data sources. Research in this area has progressed through three stages: teaching individual models to use tools, standardising how tools are discovered and invoked, and standardising how tool-using agents communicate with each other.

\subsubsection{Tool-Augmented Language Models}

Early work focused on enabling LLMs to call tools at all. \cite{schick_toolformer_2023} demonstrated that a 6.7B-parameter model (GPT-J) can teach itself to use tools through a self-supervised pipeline: candidate API calls are sampled at each position in the training text, executed, and retained only if including the result reduces the cross-entropy loss over subsequent tokens. Trained on five tools (calculator, QA system, Wikipedia search, machine translation, calendar), the model selected the correct tool in 97.9\,\% of cases on mathematics benchmarks and outperformed GPT-3 (175B) despite being 25$\times$ smaller---but is limited to a single tool call per example and cannot reformulate queries when results are unhelpful.

\cite{patil_gorilla_2023} addressed the tool-selection problem at larger scale by fine-tuning LLaMA-7B on 16{,}450 instruction--API pairs across 1{,}645 APIs from three model hubs. On their APIBench benchmark, Gorilla achieved 20.43 percentage points higher zero-shot accuracy than GPT-4 on TorchHub and reduced hallucinated API calls---defined as invocations of entirely non-existent tools, verified by AST sub-tree matching against the API database---from 36.55\,\% (GPT-4) to 6.98\,\%. With a ground-truth retriever, hallucination dropped to 0\,\% on TorchHub (7.08\,\% on HuggingFace, 1.89\,\% on TensorFlow Hub).

\cite{qu_tool_2024} synthesised these findings into a four-stage workflow---task planning, tool selection, tool calling, and response generation---and surveyed benchmarks scaling up to 16{,}464 tools (ToolBench). Their central finding is that putting all tool descriptions into the LLM's context becomes impractical beyond a few hundred tools due to length and latency constraints, making a retrieval-based pre-filtering stage mandatory. The implication for multi-domain systems is direct: Training Copilot's design of scoping each specialist to at most two MCP servers (1--26 tools per specialist, depending on domain) follows this principle, keeping tool selection well below the threshold where retrieval becomes mandatory.

\subsubsection{From Ad-hoc Integration to Standardised Protocols}

The proliferation of tool-augmented agents created a standardisation problem: each application implemented its own integration, producing an $M \times N$ fragmentation of $M$ applications connecting to $N$ data sources. Two complementary protocols address this at different layers.

The \textbf{Model Context Protocol (MCP)}~\citep{anthropic_mcp_2024, anthropic_mcp_announcement_2024} defines a universal client--server protocol based on JSON-RPC\,2.0~\citep{jsonrpc_spec_2013} for tool discovery and invocation. It provides three primitives---Prompts, Resources, and Tools---and critically separates authentication from tool declaration, so a single host can discover and use tools from arbitrary servers without provider-specific code. \cite{jayanti_enhancing_2026} proposed Context-Aware MCP (CA-MCP), which adds a \textit{Shared Context Store} (SCS) accessible by all servers: the central LLM seeds the store with goals and constraints, after which servers self-coordinate by reading and writing to it---reducing LLM orchestration calls by 67.8\,\% on the TravelPlanner benchmark while improving task completeness. Training Copilot does not adopt the SCS pattern (our A2A layer handles inter-agent coordination instead), but the CA-MCP results validate MCP's extensibility as a protocol foundation.

The \textbf{Agent-to-Agent (A2A) protocol}~\citep{google_a2a_2025}, donated to the Linux Foundation after its April 2025 announcement by Google, provides a JSON-RPC\,2.0-based specification for direct inter-agent communication. Its key design choice is \textit{opaque internals}: agents advertise capabilities via Agent Cards but need not expose their reasoning, only their results---enabling composition of agents built with different frameworks and providers. \cite{yang_survey_2025} surveyed the protocol landscape and drew a useful distinction: MCP is \textit{context-oriented} (connecting agents to data), while A2A is \textit{coordination-oriented} (connecting agents to each other). Together they address both halves of the integration problem, but to date no published system has combined both protocols in a domain-specific multi-agent application.

\subsection{Sports Analytics and Wearable Technology}\label{sec:sota:sports}

Sports science literature converges on a central finding relevant to this work: effective training management requires integrating multiple physiological and contextual signals, yet no consumer tool performs this integration with reasoning. We organise the evidence around three themes: training load and periodisation, recovery monitoring, and wearable data reliability.

\subsubsection{Training Load: Monitoring and Injury Prevention}

\cite{bourdon_monitoring_2017} published an international consensus statement (11 authors from five countries) formalising the distinction between \textit{external load}---objective measures of work performed (distance, speed, acceleration, power output, measured by GPS and accelerometers)---and \textit{internal load}---the psychophysiological response (heart rate, HRV, session RPE, blood lactate). Their central recommendation is that both must be monitored in combination, because the same external load produces different internal responses depending on fatigue, illness, or emotional state; this ``uncoupling'' is itself a diagnostic signal. They also establish the acute-to-chronic workload ratio (ACWR) as a key metric, with 0.8--1.3 as the low-risk range. \cite{halson_monitoring_2014} complemented this from the fatigue side, concluding that no single biomarker is sufficiently validated on its own.

\cite{gabbett_training_2016} formalised the ``training--injury prevention paradox'' using data from cricket, Australian football, and rugby league: athletes with well-developed chronic training loads have \textit{fewer} injuries than undertrained athletes, because physical conditioning is itself protective. The critical metric is the ACWR: at ratios $\leq$0.99, injury likelihood in the subsequent seven days was approximately 4\,\% (in elite cricket fast bowlers); at ratios $\geq$1.5, this risk increased 2--4$\times$. Week-to-week load increases of 15\,\% or more were associated with injury rates of 21--49\,\%, versus below 10\,\% when increases stayed within 5--10\,\%. \cite{grgic_periodization_2017} extended this to periodisation, finding in a meta-analysis that linear and undulating periodisation produce equivalent hypertrophy outcomes, suggesting that the specific periodisation pattern matters less than the principle of structured load variation itself. Together, these findings establish that effective load management requires \textit{trend analysis over time}---the ACWR, rolling averages, monotony indices---not point-in-time metrics. Training Copilot's load specialist computes CTL, ATL, and TSB---derived from the fitness--fatigue impulse-response model originally proposed by \cite{morton_modeling_1990}---from Strava data to operationalise these trends. We note that the ACWR itself has faced methodological criticism: \cite{impellizzeri_training_2020} identified mathematical coupling in the ratio (the acute load appears in both numerator and denominator) and showed that simulation studies can produce spurious ACWR--injury associations from data with no true relationship; our system therefore presents ACWR as one contextual signal among several rather than a standalone injury predictor.

\subsubsection{Recovery: From Single Metrics to Multi-Signal Assessment}

Recovery monitoring has shifted from single-metric approaches toward multi-signal integration. \cite{kellmann_recovery_2018} published a consensus statement arguing that systematic recovery monitoring---not just load tracking---should be integral to any training programme, as the stress--recovery balance determines long-term performance sustainability.

Heart rate variability (HRV) has emerged as a central biomarker, but its interpretation is not straightforward. \cite{plews_training_2013} showed, using data from Olympic and World Champion rowers, that in elite athletes both increases \textit{and} decreases in HRV can indicate negative adaptation---because at very low resting heart rates, vagal HRV saturates (the relationship between HRV and R-R interval is quadratic, not linear). They recommend using weekly rolling averages of Ln\,rMSSD (not single-day values, which showed no meaningful correlation with performance: $r = -0.17$) and interpreting each athlete's values against their individual baseline rather than population norms. Among four gold-medal-winning rowers, the correlation between Ln\,rMSSD and R-R interval ranged from $r = 0.91$ to $r = -0.03$---fundamentally different physiological profiles that would be misinterpreted by any system using fixed thresholds. \cite{lundstrom_hrv_2023} reviewed the practical implementation of HRV-guided training decisions in consumer settings, identifying the gap between clinical-grade and consumer-grade HRV measurement.

The strongest evidence for multi-signal approaches comes from \cite{rothschild_predicting_2024}, who tracked 43 endurance athletes (67\,\% triathletes) over 12 weeks (3{,}572 person-days), collecting daily training load, dietary intake, sleep, subjective well-being, and resting HRV. They tested nine ML algorithms (Ridge, LASSO, SVM, XGBoost, LightGBM, among others) to predict daily perceived recovery status (PRS). The best group model (LASSO) achieved $R^2 = 0.31$ (RMSE 11.8 on a 0--100 scale); individual models using each athlete's top-5 features reached $R^2 = 0.35 \pm 0.20$, with RMSE varying more than fivefold across participants (5.5--23.6). The most predictive features were soreness, sleep index, and prior-day PRS---not HRV alone. Their conclusion is that recovery prediction requires multi-metric monitoring, but the optimal feature set is individual-specific. \cite{wackerhage_personalized_2021} situated this in a broader argument: a training plan is inherently a complex mix of interacting interventions that change over time, making it virtually impossible to base every decision on scientific evidence alone. This complexity motivates tool-assisted approaches: Training Copilot's recovery specialist simultaneously queries HRV, sleep, Body Battery, and stress data to approximate the multi-signal assessment these studies advocate.

\subsubsection{Wearable Reliability and Data Ecosystem Biases}

The reliability of consumer wearable data is a prerequisite for any analytics system built upon it. \cite{fuller_reliability_2020} systematically reviewed 158 publications covering 45 device models from nine manufacturers (including Garmin, Fitbit, Apple Watch, Polar, and Samsung). For step counting under controlled conditions, 45\,\% of 979 comparisons fell within $\pm$3\,\% of criterion measures, with an overall tendency to underestimate (median error: $-$2\,\%). For heart rate, 57\,\% of comparisons were within $\pm$3\,\% error (Apple Watch: 71\,\%; Fitbit: 51\,\%; Garmin: 49\,\%). Precision degraded under two conditions: during high-intensity exercise (heart-rate correlation dropped from very strong to moderate) and in free-living settings (error dispersion widened for all metrics). Energy expenditure was inaccurate across all devices. Interdevice reliability for steps was very strong (35 of 38 correlation coefficients rated very strong), but intradevice reliability was variable (mean $r = 0.58$). \cite{seckin_review_2023} highlighted that translating raw sensor data into actionable coaching insights remains an open challenge. \cite{vanhooren_wearables_2020} reviewed real-time wearable feedback in running, identifying challenges around feedback timing and injury risk communication.

Data ecosystem biases further complicate interpretation. \cite{venter_bias_2023} quantified selection bias in Strava's crowdsourced data, finding overrepresentation of recreationally active users, males, and middle-aged adults (35--54). While professional adoption of wearable monitoring is now standard in elite team sports~\citep{bourdon_monitoring_2017}, the transition to consumer athletes introduces noise and bias that any analytical system must acknowledge. Training Copilot mitigates this partially through multi-source corroboration---cross-referencing HRV, sleep, and Body Battery rather than relying on any single metric---but cannot fully compensate for sensor inaccuracies.

Across all three themes, a consistent pattern emerges: sports science has established that training decisions require integrating load trends, recovery signals, and contextual factors, yet the consumer tools that \textit{collect} this data do not \textit{reason across} it. The gap is not in data availability but in cross-domain synthesis.

\subsection{Retrieval-Augmented Generation}\label{sec:sota:rag}

Retrieval-Augmented Generation combines the factual precision of information retrieval with the fluency of generative models. \cite{lewis_retrieval-augmented_2020} introduced the foundational architecture: a BART-large generator (400M parameters) augmented with a Dense Passage Retriever (DPR) that queries a FAISS index of 21 million Wikipedia passages at inference time. They proposed two variants: RAG-Sequence, which uses a single retrieved passage to condition the entire output, and RAG-Token, which can draw a different passage for each output token. RAG set new state-of-the-art on three open-domain QA benchmarks: 44.5 EM on Natural Questions (RAG-Sequence), 45.5 on WebQuestions (RAG-Token), and 52.2 on CuratedTrec (RAG-Sequence). In human evaluation on question generation, RAG was judged more factual than BART in 42.7\,\% of comparisons versus 7.1\,\% for BART, and produced more diverse outputs (distinct tri-gram ratio: 53.8\,\% vs.\ 32.4\,\%). A key practical property is that the retrieval index can be hot-swapped without retraining to update world knowledge.

The retrieval component builds on \cite{karpukhin_dense_2020}, who demonstrated that a dual-encoder framework (DPR) outperforms BM25 by 9--19\,\% in top-20 retrieval accuracy on multiple benchmarks. \cite{reimers_sentence-bert_2019} proposed Sentence-BERT (SBERT), which adapts the BERT architecture through a Siamese network training objective so that the resulting fixed-size vectors can be compared efficiently via cosine similarity---the \texttt{sentence-transformers} library built on this work provides Training Copilot's embedding model (\texttt{all-MiniLM-L6-v2}, $\sim$90\,MB, running locally without an embedding API). \cite{gao_retrieval-augmented_2024} surveyed the progression from Na\"ive RAG (retrieve-then-generate) through Advanced RAG (with query rewriting and re-ranking) to Modular RAG (pluggable components). Training Copilot implements a Na\"ive RAG approach: embed, retrieve by cosine similarity, generate with retrieved context---which is sufficient for its curated corpus of $\sim$3{,}000 chunks from eight public-domain books.

\subsection{AI-Assisted Sports Coaching}\label{sec:sota:coaching}

The application of AI to sports coaching is an emerging field where three limitations recur across studies: insufficient access to the athlete's actual data, limited cross-domain reasoning, and single-agent architectures that cannot scale to heterogeneous tool sets.

\textbf{The data-access and personalisation gap.} \cite{monteiro-guerra_personalization_2020} reviewed 28 papers covering 17 real-time physical-activity coaching applications and classified them along seven personalisation dimensions. Feedback (17/17 apps) and goal setting (15/17) were universal, but the more sophisticated features were rare: context-awareness appeared in only 3 of 17 systems, and adaptation (adjusting behaviour based on key behavioural factors) in only 2 of 17. Of the three context-aware systems, only one provided implementation details. \cite{luo_promoting_2021} reached a similar conclusion for conversational agents specifically: chatbots show promise for sustained engagement, but most operate without access to wearable data. \cite{puce_harnessing_2025} confirmed this pattern in a systematic review of 10 studies on generative AI for exercise prescription, finding that AI-generated plans adhere to foundational training principles but lack specificity, progression, and---most critically---the ability to incorporate real-time physiological feedback. Plans did not adapt to individual context, and identical prompts sometimes produced different outputs, raising reproducibility concerns.

\textbf{Single-agent and single-domain architectures.} The most directly relevant prior system is GPTCoach~\citep{joerke_gptcoach_2024}, an LLM-based coaching chatbot evaluated at CHI~2025. GPTCoach uses GPT-4 with a prompt-chaining architecture (separate LLM calls for dialogue-state classification, motivational-interviewing strategy selection, and tool-call decisions) and two tools that query Apple HealthKit data (three months of step count, heart rate, active energy, and other metrics via HL7 FHIR). In a lab study with 16 participants, it scored 4.8/5 on feeling supported and 4.5/5 on empathy; external MITI coding found 93\,\% MI-consistent behaviour. However, GPTCoach operates as a \textit{single} agent with only two tools and accesses data from a single source (HealthKit). It does not integrate weather, calendar, route planning, or cross-domain synthesis---and its evaluation was limited to a single onboarding session, not longitudinal coaching.

At the architectural level, \cite{vemuri_multiagent_2021} proposed SMART-ALEC, a multi-agent health coaching system with six internal modules per user agent (planner, scheduler, learner, context analyser, executor, communicator) for physical activity promotion. They implemented two prototypes: BeSmart (SMS-based, beta-tested with 250 agents) and a JIT intervention system using Apple Watch with nightly-updated decision trees. While predating MCP and A2A and using ad-hoc coordination (GPGP-style commitment exchange), their design philosophy---modular specialists with a central coordinator, transferable learned policies between similar users---validates the multi-agent approach. Training Copilot extends this with standardised protocols (MCP for tool access, A2A for coordination) and LLM-based reasoning rather than finite-state-machine dialogues.

\textbf{The human-AI boundary.} \cite{barger_artificial_2025} conducted a mixed-methods RCT ($N = 52$) comparing coaching sessions where participants believed they were coached by an AI (actually a human coach behind a live-motion avatar with voice distortion) versus a human coach. Working alliance scores were not significantly different (AI: $M = 72.73$, human: $M = 74.50$; $t(50) = -0.71$, $p = 0.48$). Qualitatively, participants in the AI condition reported using fewer impression-management behaviours and greater willingness to disclose sensitive topics. The authors recommend a synergy model: AI coaches designed with expert coaching models, narrowly focused on specific areas, working alongside human coaches. \cite{gabarron_human_2024} reviewed AI-driven solutions for physical activity more broadly, finding recommender systems most studied, followed by conversational agents, but noting sparse long-term evidence for both.

In summary, prior coaching systems either have data access (wearable-native apps like GPTCoach, limited to a single source) or reasoning capability (general-purpose LLMs, without live data), but none combine multi-source data integration, cross-domain reasoning, and grounding in verifiable literature via standardised protocols.

\subsection{Summary and Research Gap}\label{sec:sota:summary}

The literature reviewed above reveals a consistent pattern: each research stream has matured to the point where its individual contribution is well-established, yet all share a common limitation---\textit{isolation from the complementary capabilities needed for integrated training support}.

\textbf{Agentic AI} has demonstrated that multi-agent architectures with domain-scoped specialists outperform monolithic designs (Section~\ref{sec:sota:agents}), but existing systems rely on ad-hoc tool registries and custom communication layers rather than standardised protocols, limiting their extensibility and reproducibility.

\textbf{Tool integration protocols} have solved the $M \times N$ fragmentation problem in principle (Section~\ref{sec:sota:protocols}): MCP standardises tool discovery and invocation, while A2A standardises inter-agent coordination. However, no published system has combined both protocols in a domain-specific multi-agent application---they remain demonstrated in generic or single-domain contexts.

\textbf{Sports science} has established that training decisions require integrating load trends, recovery signals, and contextual factors (Section~\ref{sec:sota:sports}), with consensus statements~\citep{bourdon_monitoring_2017, kellmann_recovery_2018} and empirical studies~\citep{rothschild_predicting_2024} showing that no single metric suffices. Yet consumer platforms \textit{display} these metrics in isolation without \textit{reasoning across} them.

\textbf{AI-assisted coaching} systems have shown that LLM-based coaches can produce guidance perceived as personalised and actionable~\citep{joerke_gptcoach_2024}, but existing systems either access data from a single source (wearable-native apps) or reason without real data access (general-purpose LLMs)---none combine multi-source data integration, cross-domain reasoning, and grounding in verifiable literature (Section~\ref{sec:sota:coaching}).

\textbf{RAG} enables grounding in verifiable sources (Section~\ref{sec:sota:rag}), addressing the reliability concerns raised by \cite{puce_harnessing_2025}, but has not been applied to sports coaching in combination with live physiological data.

The research gap is therefore not in any individual capability but in their \textit{integration}: no existing system combines standardised multi-source data access (MCP), multi-agent coordination (A2A), domain-scoped specialist reasoning (LangGraph ReAct), and retrieval-grounded knowledge (RAG) into a platform for athlete-facing training support. Training Copilot addresses this gap directly: its architecture maps each identified limitation to a specific design decision, as detailed in Section~\ref{sec:architecture}.
