\markboth{SYSTEM ARCHITECTURE}{SYSTEM ARCHITECTURE}
\section{System Architecture}\label{sec:architecture}

This section presents Training Copilot's three-layer architecture: the MCP data layer, the agent layer, and the LLM integration seam. The guiding principle is \textit{tool-agnostic extensibility}: no component names a specific tool or data source. Tools are discovered at runtime, servers are registered declaratively, and the LLM provider is resolved from configuration---so the entire stack can be extended without code changes.

\subsection{Design Principles}\label{sec:architecture:principles}

The architecture adheres to six principles derived from the MCP specification and microservice design patterns~\citep{dragoni_microservices_2017}: (1)~\textbf{tool-agnostic}---the model discovers tools via \texttt{list\_tools()} and decides autonomously which to call; (2)~\textbf{one call path}---\texttt{ToolHost.call\_tool()} is the single entry point for every invocation; (3)~\textbf{servers as independent services}---each MCP server is a self-contained process with its own port, authentication, and lifecycle; (4)~\textbf{discovery over hardcoding}---unreachable servers are silently skipped; (5)~\textbf{auth separated from declaration}---credentials are passed as connection headers, never exposed to the model; (6)~\textbf{vendor-neutral LLM seam}---switching providers is a configuration change.

\subsection{Design Rationale}\label{sec:architecture:rationale}

Each major architectural decision maps directly to a limitation identified in the state of the art (Section~\ref{sec:sota:summary}):

\textbf{Why MCP rather than direct API integration?} Sports science consensus statements~\citep{bourdon_monitoring_2017, kellmann_recovery_2018} establish that training decisions require data from multiple heterogeneous sources. Direct API integration would create the $M \times N$ fragmentation that \cite{yang_survey_2025} identified as the core scalability barrier for tool-augmented agents. MCP's standardised discovery and invocation protocol reduces this to $M + N$: each new data source requires one server and one configuration line, with no changes to the host, agents, or UI.

\textbf{Why five specialist agents rather than one?} \cite{qu_tool_2024} and \cite{patil_gorilla_2023} showed that tool selection accuracy degrades as the tool set grows. Our early prototyping confirmed this: a single agent with access to all 40+ tools frequently selected incorrect tools or conflated domains (e.g., using weather data to answer recovery questions). Scoping each specialist to at most two MCP servers restores tool selection accuracy while enabling domain-specific prompts---the recovery specialist knows how to interpret HRV relative to baseline, the route specialist knows to geocode before routing. The five-domain decomposition (recovery, load, context, routes, fitness knowledge) follows the established sports science distinction between internal load, external load, environmental factors, route planning, and evidence-based guidance~\citep{bourdon_monitoring_2017, gabbett_training_2016, kellmann_recovery_2018}.

\textbf{Why LangGraph for agent execution?} The specialist agents need a framework that supports the ReAct loop~\citep{yao_react_2023}---interleaved reasoning and tool calling---while allowing the orchestrator to treat each specialist as an opaque unit. LangGraph~\citep{langchain_langgraph_2024} provides this through its graph-based execution model: each agent is a state machine where nodes represent reasoning or action steps and edges encode the control flow. Compared to alternatives such as AutoGen~\citep{wu_autogen_2023} (conversation-pattern-based, less suited to independent specialist processes) or custom prompt loops (no built-in state management or tracing), LangGraph offers three concrete advantages: (1)~native integration with LangChain's tool abstraction, enabling our MCP-to-LangChain bridge to expose MCP tools as standard LangChain tools without adapter code; (2)~built-in support for MLflow autologging, which provides the hierarchical span trees the evaluation framework depends on (Section~\ref{sec:evaluation:trace}); and (3)~a clear process boundary per agent, allowing each specialist to run as an independent A2A server with its own scoped ToolHost.

\textbf{Why A2A for inter-agent communication?} The orchestrator--specialist pattern requires a protocol that supports capability advertisement, task delegation, and result collection without coupling agents to each other's internals. A2A~\citep{google_a2a_2025} provides exactly this: Agent Cards advertise capabilities, task-based delegation returns structured artifacts, and opaque internals mean specialists can be replaced or extended independently. Using A2A over direct function calls also enables the deployment flexibility the system requires---agents can run in-process, as separate local processes, or in separate containers, with only URL changes.

\subsection{Architecture Overview}\label{sec:architecture:overview}

Figure~\ref{fig:architecture} shows the complete system architecture. The data path flows from the UI through the agent layer to MCP servers and back; the chat path adds A2A coordination on top.

\input{1_Figures/fig_architecture}

\subsection{Layer 1: The MCP Data Layer}\label{sec:architecture:mcp}

The foundation is the Model Context Protocol~\citep{anthropic_mcp_2024}, a JSON-RPC\,2.0-based~\citep{jsonrpc_spec_2013} client--server protocol where a \textit{host} discovers tools from one or more \textit{servers} and routes invocations transparently. Training Copilot uses the Streamable HTTP transport (stateless, one request per tool call), so each server is an independent HTTP endpoint and requires no persistent connection.

The server registry is a declarative mapping from logical names to Streamable HTTP endpoints in a single file (\texttt{core/config.py}, Listing~\ref{lst:registry} in the Appendix). Each URL is constructed via a helper that reads an environment variable (e.g., \texttt{WEATHER\_MCP\_URL}), falling back to a localhost default. This means switching from local processes to Docker Compose containers is a URL change, not a code change. Adding a new data source requires one line in this dictionary and one server file.

The \texttt{ToolHost} class implements two operations. \texttt{list\_tools()} opens a fresh MCP session to every registered server \textit{in parallel} via \texttt{asyncio.gather}, calls \texttt{list\_tools()} on each session, and namespaces the returned tools as \texttt{server\_\_tool} (the double-underscore separator is necessary because dots are not legal in OpenAI function-call names). Each tool's JSON Schema parameters and its docstring are preserved verbatim---the docstring is the only text the LLM reads when deciding whether and how to call a tool. Servers that fail to connect within the 60-second timeout are silently skipped, returning an empty list rather than raising an exception. This is the mechanism behind graceful degradation: the system functions with whichever servers are reachable.

\texttt{call\_tool(name, args)} splits the namespaced name on the separator, looks up the server URL, opens a session, calls the tool, and returns the result as a JSON string. Tool errors (including timeouts and connection failures) are returned as structured \texttt{\{"error": "..."\}} objects rather than exceptions, so agents can reason about failures and report them to the user rather than crashing. The async implementation uses one fresh event loop per synchronous call via a bridge function, avoiding deadlocks when called from thread-pool workers (e.g., the Streamlit runtime).

Each MCP server is a self-contained FastMCP service (Listing~\ref{lst:server_pattern} in the Appendix). Per-server credentials are passed as connection headers via the \texttt{ToolHost(headers=\{...\})} constructor---the authentication layer is separate from the tool declaration, so credentials never enter the model's context and tool definitions remain portable across deployments.

\subsection{Layer 2: The Agent Layer}\label{sec:architecture:agents}

The agent layer combines LangGraph~\citep{langchain_langgraph_2024} for within-agent execution with A2A~\citep{google_a2a_2025} for between-agent coordination. Each of the five specialist agents is a LangGraph ReAct agent~\citep{yao_react_2023} with access to a \textit{scoped} ToolHost that discovers tools only from its assigned MCP servers (Table~\ref{tab:agent_scope}). This scoping is enforced at the infrastructure level---a specialist physically cannot call tools outside its domain---rather than relying on prompt instructions that the model might ignore under complex queries.

\begin{table}[ht]
\begin{center}
\begin{minipage}{\textwidth}
\caption[Agent--MCP scope mapping]{Agent-to-MCP-server scope mapping. Each specialist discovers tools only from its assigned servers.}\label{tab:agent_scope}
\begin{tabularx}{\textwidth}{llX}
\toprule
\textbf{Agent} & \textbf{MCP Servers} & \textbf{Domain} \\
\midrule
Recovery (:9001) & Garmin & Sleep, HRV, Body Battery, stress, readiness \\
Load (:9002) & Strava, Garmin & Training load (CTL/ATL/TSB), volume, splits, zones \\
Context (:9003) & Weather, Calendar & Forecast, UV, pollen; calendar read/write \\
Route (:9004) & Routes & Route planning, geocoding, trail search, isochrones \\
Fitness (:9005) & \textit{(none---RAG)} & Exercise science from a vector DB of literature \\
\botrule
\end{tabularx}
\end{minipage}
\end{center}
\end{table}

The bridge between MCP tools and LangGraph agents is \texttt{core/mcp\_langchain.py}. \texttt{scoped\_host(server\_names)} constructs a \texttt{ToolHost} restricted to a subset of the registry---a specialist physically cannot discover tools outside its domain, regardless of what its prompt says. \texttt{build\_tools(host, recorder)} then calls \texttt{alist\_tools()} on the scoped host and wraps each discovered tool as a LangChain \texttt{StructuredTool}. The wrapper implements the record-full/clip-for-model pattern (Section~\ref{sec:agents:recording}): the full JSON result is appended to a shared \texttt{recorder} list (which becomes the A2A DataPart artifact), while the LLM receives a truncated copy where large arrays (GPS points, intraday timelines) are replaced with placeholders such as \texttt{[847 items]} and the total string is capped at 6{,}000 characters.

The orchestrator agent (port 9000) is a LangGraph agent whose only tools are \texttt{ask\_<specialist>} functions. Each is a thin wrapper around \texttt{core/a2a\_client.call\_agent()}, which resolves the specialist's Agent Card (at \texttt{/.well-known/agent-card.json}), opens a streaming A2A session via \texttt{SendStreamingMessageRequest}, and demuxes the response into three streams: status-update messages (relayed as progress strings to the UI), text artifact parts (the answer), and data artifact parts (the specialist's full tool-call records). The orchestrator has no MCP access of its own: it decomposes the query, delegates to specialists in parallel when domains are independent (the LLM emits multiple \texttt{ask\_*} calls in a single step, which LangGraph executes concurrently), synthesises their responses, and assembles the \texttt{trace} structure the UI uses for maps, charts, and the debug panel.

Agents run non-streaming internally (\texttt{ainvoke}, not \texttt{astream}) for robustness against the KIT university gateway, which has exhibited instabilities under streaming load. Progress is surfaced through A2A status messages instead of token-level SSE.

\subsection{Layer 3: The LLM Integration Seam}\label{sec:architecture:llm}

The LLM seam (\texttt{core/llm.py}) is a vendor-neutral abstraction supporting three backends: OpenAI-compatible endpoints (including the KIT university gateway), official OpenAI, and Google Gemini via \texttt{ChatGoogleGenerativeAI}. The seam re-reads \texttt{.env} per call, so provider changes via the Settings UI take effect without restarting agent processes.

\input{1_Figures/fig_placeholder_chat}

\input{1_Figures/fig_placeholder_dashboards}

\subsection{Supporting Subsystems}\label{sec:architecture:supporting}

\textbf{Per-user memory.} Conversation history is embedded with the same local \texttt{all-MiniLM-L6-v2} model as the fitness RAG pipeline~\citep{reimers_sentence-bert_2019} and searched by cosine similarity to inject relevant prior exchanges into the prompt. A background distillation thread periodically condenses the history into a human-readable profile (\texttt{soul.md}) of durable facts---training goals, injury history, schedule constraints---prepended to every subsequent prompt. Both layers degrade gracefully if unavailable.

\textbf{LLM-driven chart generation.} Given tool results from a completed turn, the LLM generates Plotly Python code executed server-side in a restricted namespace. If execution fails, a single Reflexion-style retry~\citep{shinn_reflexion_2023} is attempted; still-failing charts are dropped. This avoids hardcoded chart types: the LLM can produce any visualisation the data supports.

\textbf{Multi-channel delivery.} The orchestrator adapter exposes a uniform \texttt{run()} interface consumed by three frontends: a React app backed by FastAPI SSE, a Telegram userbot bridge (with voice-memo transcription and portable route formats), and the original Streamlit chat tab. All share the same agent layer; only rendering differs.

\textbf{3D activity flythrough.} A cinematic camera animation along GPS tracks over satellite basemaps, with bearing, pitch, and zoom EMA-smoothed for cinematic fluidity. In-browser WebCodecs export with hardware-acceleration probing; a headless Chromium path via Playwright for server-side MP4 rendering.

\input{1_Figures/fig_placeholder_flythrough}

\input{1_Figures/fig_placeholder_sync}

\subsection{Deployment}\label{sec:architecture:deployment}

Training Copilot supports local development (each service as a separate process, started by \texttt{dev\_stack.sh}), Docker Compose (single Dockerfile reused per service, URLs swapped via env vars), and hybrid modes. The declarative registry means switching between modes is a URL change, not a code change.
