\markboth{DATA SOURCES AND INTEGRATION}{DATA SOURCES AND INTEGRATION}
\section{Data Sources and Integration}\label{sec:data}

Training Copilot integrates seven data sources through the uniform MCP interface described in Section~\ref{sec:architecture:mcp}. Each source is encapsulated in an independent server handling authentication, rate limiting, error recovery, and data normalisation internally. Table~\ref{tab:data_summary} summarises the sources, tool counts, and metrics. This section describes the integration considerations for each.

\subsection{Strava --- Activity Tracking}\label{sec:data:strava}

The Strava MCP server (port 8103) connects to the Strava v3 REST API. Authentication follows the standard OAuth\,2.0 Authorization Code flow: the developer registers an application on the Strava Developer Portal, obtaining a \texttt{CLIENT\_ID} and \texttt{CLIENT\_SECRET}. On first use, the server opens the user's browser to Strava's consent page, where the user grants the requested scopes (\texttt{read}, \texttt{activity:read\_all}, \texttt{activity:write}). Strava redirects to a local callback server (\texttt{localhost:8080/callback}) with an authorization code. The server exchanges this code for an access token and a refresh token, persists both to \texttt{.tokens/strava.json}, and automatically refreshes the access token using the refresh token before each API call, with a five-minute buffer before the server-reported expiry time to prevent edge-case expiry during a request. The entire flow is implemented in \texttt{auth/strava\_oauth.py} and is transparent to the MCP tools---they call the API with a valid token and do not handle authentication.

The server's thirteen tools cover recent activities, aggregate statistics, year-over-year breakdowns, personal records, gear mileage, deep single-activity detail with lap splits and power data, GPS streams at sub-second resolution, performance trend analysis against personal baselines, and the Training Stress Balance metrics (CTL, ATL, TSB) derived from the fitness--fatigue impulse-response model~\citep{bourdon_monitoring_2017}. HTTP errors are handled with exponential backoff for server errors (5xx) and immediate return for rate-limit responses (429), since Strava's \texttt{Retry-After} is typically 15+ minutes, making in-request retries pointless.

In June 2026, Strava announced that API access now requires a Strava subscription for Standard Tier developers, alongside the launch of an official Strava MCP server for AI-native data access~\citep{strava_api_update_2026}. This development validates Training Copilot's MCP-first architecture: when Strava's own MCP server matures, it could replace our custom server with a URL change in the registry, no agent or UI code changes.

\subsection{Garmin Connect --- Wellness and Recovery}\label{sec:data:garmin}

Unlike Strava, Garmin does not provide a public REST API for the health and wellness data (sleep, HRV, Body Battery, stress) that Training Copilot needs. The Garmin Health API is limited to enterprise partners. Our server therefore relies on the open-source \texttt{garminconnect} Python library,\footnote{\url{https://github.com/cyberjunky/python-garminconnect}} which reverse-engineers the Garmin Connect web application's internal endpoints. Authentication proceeds through Garmin's SSO login page using stored email and password credentials, with MFA (one-time passcode) required on first login. The library caches session tokens in \texttt{.tokens/}; subsequent logins reuse cached tokens and only re-authenticate when they expire.

This approach carries inherent risks. The internal API endpoints are undocumented and may change without notice, potentially breaking the server. The \texttt{garminconnect} library is a community project with no official support or stability guarantees. Rate limits are stricter and less predictable than Strava's: we observed that the Body Battery endpoint rejects date ranges longer than 28 days, so the server automatically chunks requests. A retry mechanism with exponential backoff handles transient failures (timeouts, 429s, 502/503 errors). A mock-data mode (\texttt{GARMIN\_MOCK\_HEALTH=true}) enables UI development without a connected device.

The server's thirteen tools span seven metric categories: sleep stages with SpO\textsubscript{2} and sleep-associated HRV; last-night HRV with personal baseline and readiness status; Body Battery as daily highs, lows, and intraday timeline; intraday stress and heart rate timelines; VO\textsubscript{2}max, training load, and race predictions; body composition over configurable date ranges; and multi-day wellness trends with step timelines and activity GPS tracks. These metrics collectively enable the recovery specialist to assess training readiness through simultaneous consideration of sleep quality, HRV relative to baseline, stress, and remaining energy---the multi-signal approach that \cite{rothschild_predicting_2024} showed is necessary for accurate recovery prediction.

\subsection{Weather, Routes, and Calendar}\label{sec:data:other}

The \textbf{weather server} (port 8101) queries the Open-Meteo API, which requires no API key and imposes no rate limits for non-commercial use. Its four tools provide current conditions, multi-day forecasts (1--16 days, hourly or daily resolution) with temperature, precipitation probability, wind speed, and UV index, pollen concentrations by species, and UV category classification. The context specialist applies heuristic rules to classify training windows: 5--20\,\textdegree C is flagged as ideal temperature, precipitation probability above 30\,\% triggers a rain warning, and UV index above 6 triggers a sun-protection advisory.

The \textbf{routes server} (port 8102) aggregates three external services behind its seven tools. OpenRouteService (authenticated via \texttt{ORS\_API\_KEY}) provides A-to-B routing with multiple profiles (cycling, running, hiking), circular route generation from a start point with a target distance, elevation profiles, and isochrone maps (reachability polygons for ``what can I reach in 30 minutes?''). Google Geocoding resolves place names to coordinates. OSM Overpass queries retrieve boundary polygons for named areas, enabling park-constrained loops (e.g., ``a run inside Schlossgarten'' generates the park boundary from OpenStreetMap, then routes within it). Each tool returns GeoJSON-compatible coordinate arrays that the UI renders as interactive Folium maps.

The \textbf{calendar server} (port 8105) provides six tools for Google Calendar read/write access through the Google Calendar API. It supports two authentication modes: a single-user development mode using persisted OAuth tokens in \texttt{.tokens/google.json}, and a multi-tenant mode using per-request \texttt{Authorization: Bearer} headers. The context specialist cross-references calendar free windows with weather forecasts to suggest optimal training times, and can create calendar events for planned sessions.

\subsection{Fitness Literature --- RAG Knowledge Base}\label{sec:data:rag}

The fitness specialist (port 9005) is the only agent without an MCP server. It queries a local vector database of eight public-domain exercise-science books from Project Gutenberg\footnote{\url{https://www.gutenberg.org/}} via RAG~\citep{lewis_retrieval-augmented_2020}. Embeddings are generated by the local \texttt{all-MiniLM-L6-v2} model ($\sim$90\,MB)~\citep{reimers_sentence-bert_2019}; retrieval is brute-force cosine similarity over $\sim$3k chunks---instantaneous for this corpus size. The specialist is instructed to ground every claim in retrieved passages with explicit source citations. An \texttt{ensure\_sources()} post-processing step guarantees cited references survive the orchestrator's synthesis. Figure~\ref{fig:rag_pipeline} illustrates the pipeline.

\input{1_Figures/fig_rag_pipeline}

\subsection{Telegram --- External Server Bridge}\label{sec:data:telegram}

An optional Telegram integration (port 8106) demonstrates the architecture's ability to bridge external stdio-only MCP servers onto the Streamable HTTP bus without forking their code. The upstream \texttt{chigwell/telegram-mcp} project\footnote{\url{https://github.com/chigwell/telegram-mcp}} (116 tools) is run unmodified via \texttt{uv} in an isolated environment; to the \texttt{ToolHost}, it looks identical to every other server---confirming the vendor-agnostic promise.

\subsection{Summary}\label{sec:data:summary}

\begin{table}[ht]
\begin{center}
\begin{minipage}{\textwidth}
\caption[Summary of integrated data sources]{Summary of integrated data sources.}\label{tab:data_summary}
\begin{tabularx}{\textwidth}{llcX}
\toprule
\textbf{Source} & \textbf{Server} & \textbf{Tools} & \textbf{Primary Metrics} \\
\midrule
Strava & :8103 & 13 & Activities, load, trends, splits, GPS, records, analysis \\
Garmin & :8104 & 13 & Sleep, HRV, Body Battery, stress, VO\textsubscript{2}max \\
Weather & :8101 & 4 & Forecast, UV, pollen, current conditions \\
Routes & :8102 & 7 & A-B routes, loops, trails, isochrones, elevation \\
Calendar & :8105 & 6 & Events, free slots, event detail, create/update/delete \\
Literature & RAG & 1 & Exercise-science passages (8 books) \\
Telegram & :8106 & 116 & Messages, chats, contacts, media \\
\botrule
\end{tabularx}
\end{minipage}
\end{center}
\end{table}
