\markboth{AGENT INTERACTION}{AGENT INTERACTION}
\section{Agent Interaction and Communication}\label{sec:agents}

This section details how the six agents coordinate to answer user queries: communication protocols, delegation patterns, prompt architecture, the recording/clipping strategy for large tool results, and observability infrastructure.

\subsection{Communication Protocols}\label{sec:agents:protocols}

Training Copilot employs two JSON-RPC\,2.0-based protocols at different layers.

\textbf{MCP} governs all data access. Each specialist calls its servers via a scoped \texttt{ToolHost} over Streamable HTTP. The scoping is enforced at the infrastructure level: when a specialist's process starts, \texttt{scoped\_host(["garmin"])} constructs a \texttt{ToolHost} that contains only the Garmin URL. The specialist's tool-discovery call returns only Garmin tools---it cannot even see the Strava or Weather tools, regardless of what its prompt says. This is a stronger guarantee than prompt-based access control, which the model can ignore under complex queries.

\textbf{A2A} governs inter-agent communication. The orchestrator calls each specialist via \texttt{ask\_<name>} tools, each performing a streaming A2A request using the \texttt{a2a-sdk} (version 0.3.x). The A2A wire format works as follows: the orchestrator's \texttt{call\_agent()} resolves the specialist's Agent Card at \texttt{/.well-known/agent-card.json}, then sends a \texttt{SendStreamingMessageRequest} containing the question. The specialist's response is a stream of typed events: \textit{status-update} events (relayed to the UI as progress messages like ``Consulting recovery agent\ldots''), \textit{artifact-update} events with \texttt{TextPart}s (the answer text), and \textit{artifact-update} events with \texttt{DataPart}s (the full tool-call records as a JSON dict). This separation is central to the architecture: the user-facing answer is concise enough for the orchestrator's LLM context, while the data artifact carries the complete MCP results (GPS streams, intraday timelines) for the UI to render maps and charts.

Figure~\ref{fig:interaction_flow} illustrates the complete flow for a representative multi-domain query.

\input{1_Figures/fig_interaction_flow}

\subsection{Delegation Patterns}\label{sec:agents:delegation}

The orchestrator employs three delegation patterns. \textbf{Single-specialist delegation} routes a query cleanly to one domain (e.g., ``What's my VO\textsubscript{2}max?'' $\to$ load specialist). \textbf{Parallel multi-specialist delegation}---the most common pattern---emits multiple \texttt{ask\_*} calls in a single LLM step, reducing latency to the duration of the slowest specialist rather than the sum. \textbf{Sequential delegation} occurs when a later specialist depends on an earlier one's output; the LLM naturally handles this across multiple reasoning steps.

\subsection{Prompt Architecture}\label{sec:agents:prompts}

The prompt system is structured as a composition of three layers, all defined in \texttt{agents/prompts.py}. This layered approach follows the principle of task-specific instruction design identified by prompt engineering research~\citep{sahoo_prompt_2024} as critical for eliciting reliable behaviour from LLMs without modifying model parameters:

A shared \textbf{base prompt} defines the agent identity, injects the current date (computed at call time, not hardcoded), sets the home location (used for geocoding defaults), and establishes five behavioural rules: (1)~use tools for any data question---never guess or fabricate; (2)~execute tool calls immediately without asking permission (``the question IS the permission''); (3)~call independent tools in parallel within a single LLM step; (4)~compute absolute dates (YYYY-MM-DD) from relative expressions before passing them to tools; and (5)~synthesise data into insight rather than dumping raw JSON.

\textbf{Specialist prompts} extend the base with a domain-specific block that enumerates the specialist's available tools with explicit \textit{when}-to-use guidance. For example, the load specialist's prompt maps ``weekly volume / consistency'' $\to$ \texttt{strava\_\_get\_training\_trends} and specifies that Strava is the primary activity source with Garmin as fallback---preventing the common failure mode where the agent queries both and produces contradictory answers from duplicated data. The recovery specialist's prompt includes interpretation rules (e.g., ``interpret HRV relative to personal baseline; flag overtraining signals when HRV is suppressed, Body Battery is low, and sleep is poor'').

The \textbf{orchestrator prompt} defines routing policy: always delegate through specialists, never answer from its own parametric knowledge, pick the minimal specialist set for each query, and preserve source citations from the fitness specialist verbatim. It does not list individual MCP tools---only specialist domains.

\subsection{Recording, Clipping, and Trace Assembly}\label{sec:agents:recording}

A recurring challenge in tool-augmented systems is that tool results can be large---a GPS stream from a two-hour ride contains thousands of lat/lon/elevation points; a 14-day wellness trend is a multi-kilobyte JSON array---while the LLM's context window is finite and each token is expensive~\citep{qu_tool_2024}. Training Copilot solves this with a \textit{record-full, clip-for-model} pattern (Figure~\ref{fig:recording_clipping}).

The implementation sits in the \texttt{build\_tools} wrapper (\texttt{core/mcp\_langchain.py}). Each tool call goes through three steps: (1)~the full JSON result is appended to a shared \texttt{recorder} list, along with the tool name, arguments, duration in milliseconds, and any error; (2)~the \texttt{clip()} function from \texttt{core/agent\_trace.py} replaces arrays under specific keys (\texttt{points}, \texttt{waypoints}, \texttt{timeline}, \texttt{buckets\_15min}, \texttt{trails}, \texttt{instructions}) with a count placeholder (e.g., \texttt{[847 items]}), truncates long string values, and caps the total string at 6{,}000 characters; (3)~the clipped string is returned to the LLM as the tool's output. The UI renders maps and charts from the \textit{full} trace, not from the model's response---decoupling data fidelity from context-window constraints.

\input{1_Figures/fig_recording_clipping}

When a specialist finishes, its accumulated \texttt{recorder} list is attached to the A2A response as a \texttt{DataPart} artifact. The orchestrator collects these artifacts from all consulted specialists and passes them to \texttt{build\_trace()}, which assembles a structured \textit{trace} dict: run ID, timestamp, original input, all tool calls with arguments, full results, duration, and error status, agents consulted, extracted route data (the first successful result from a route-producing tool), chart hints (extracted from \texttt{<!--charts: ...-->} tags the model appends), and the final answer. This trace serves three purposes: transparency (collapsible debug panel in the chat UI), rendering (route maps and inline charts from full data), and evaluation (the \texttt{grounded\_in\_real\_data} scorer in Section~\ref{sec:evaluation:scorers} reads tool spans directly from it).

\subsection{Observability and Graceful Degradation}\label{sec:agents:observability}

All agent runs are traced to an MLflow tracking server~\citep{mlflow_multiturn_2025}. Each process calls \texttt{setup\_tracing(name)} at startup, enabling LangChain autologging and wrapping each \texttt{ainvoke} in a labelled root span. The result is a hierarchical span tree per run, filterable by service, with LLM and tool calls as child spans. Tracing is strictly best-effort: a missing \texttt{mlflow} or unreachable server logs once and is ignored.

Graceful degradation is systematic at every layer: unreachable MCP servers are skipped during discovery; unavailable specialists return descriptive errors rather than causing task failure; tool errors are returned as structured JSON for agent reasoning; and MLflow tracing never blocks the critical path. This means Training Copilot provides useful answers even when some data sources or agents are offline---essential for a system depending on multiple external APIs.
